Exploring a pragmatic account of Zipf's Law of Abbreviation using Bayesian data analysis

### 2019-01-23 bda/analysis/fit_model.Rmd
beginning RD re-rog

```{r model-run-data-loading}
RUN_MODEL <- FALSE

@@

if (RUN_MODEL) {
  pt <- proc.time()
  # Baseline or Full BDA
  model_file <- getModelFileFromType(MODEL_TYPE)
  model_bda <- getModelFile(model_file)
  df_bda <- rwebppl::webppl(model_bda, data=dfAll, data_var='rData', cores=4)
  # Write file
  write.csv(df_bda, file=file.path("../analysis", paste0("df_", MODEL_TYPE, ".csv")))
  et <- proc.time() - pt
  cat(paste("Run time: ", et))
}
```

>>>>

```{r model-run-data-loading}
RUN_MODEL <- TRUE

@@

if (RUN_MODEL) {
  pt <- proc.time()
  # Baseline or Full BDA
  model_file <- getModelFileFromType(MODEL_TYPE)
  model_bda <- getModelFile(model_file)
  df_bda <- rwebppl::webppl(model_bda, data=dfAll, data_var='rData', cores=4)
  # Write file
  write.csv(df_bda, file=file.path("../analysis", paste0("df_", MODEL_TYPE, ".csv")))
  et <- proc.time() - pt
  cat(paste("Run time: ", et))
}

# Problems
# 108.196.136.202 - worked once
# 117.229.19.52 - worked once
# 122.161.26.31
# 140.141.208.86
# 80.101.212.2
# 24.113.49.186
```

### 2019-01-23 paper/paper.Rmd
update to new cogsci styleing
beginning RD re-rog

Zipf (1935) presented a view of human behavior that focused on *effort minimization* as a guiding principle. Relying on empirical evidence from language corpora, he argued that the unique dynamics of speaker- and listener-effort minimization give rise to the distributional forms found  in natural language. Among these, Zipf examined the relation between a word's frequency in a corpus and a variety of properties including its rank-frequency, its denotation size, and its length. In terms of this last property, what he called the "Law of Abbreviation," (LOA) Zipf claimed that the more frequent words are, the shorter they tend to be. Evidence for this relationship can be found across languages (CITATION), may also be present in animal communication systems  (Ferrer-i-Cancho et, 2013), and has also been observed in artificial (programming) languages (CITATION from Kanwal).\par

Zipf argued that this relationship between frequency and length was an emergent property of competing speaker-listener pressures. Under this framing, communication is viewed as a cooperative, joint activity between interlocutors. However, this joint activity contains a fundamental asymmetry -- what is effortful for a speaker (production) is different from what is effortful for a listener (understanding).  An optimal speaker language, Zipf argued, would consist of a single, low-cost word which was fully ambiguous; referring to all possible objects in the world. Adopting such a language, a listener would need to disambiguate every utterance by the speaker. By contrast, if a language were optimised solely in terms of listener effort, the language should bijectively map all forms to meanings so there was no need for a listener to disambiguate.\par

Zipf argued that these competing forces, what he called “Speaker’s economy” and “Auditor’s economy”, give rise to the particular distributional forms found in all languages, including the relationship between word frequency and length. While this framework is compelling and can largely be seen as foundational to a family of theories examining efficiency in language-structure (CITATIONS) and  -use (CITATIONS), the issue is far from settled. Subsequent work has argued that properties such as the relation between word frequency and length can theoretically be explained by a process of random typing (Ferrer-i-Cancho & Mascoso del Pardo, 2012). Kanwal et al. (2017) highlight the fact that Zipf’s original work, as well as more recent studies analysing large-scale, observational corpora, “do not explicitly target the hypothesized role of communicative pressures…”  In other words, it is difficult to make causal claims about the role that competing speaker- and listener-pressures play purely from large-scale, observational data analysis.\par

To gain traction on the causal role of speaker-listener pressures, Kanwal et al. (2017) implemented an artificial language learning paradigm to isolate the pressures directly. In line with Zipf's original analysis, they hypothesized that only when *both* speaker- and listener- pressures are present in a communication setting would the LOA emerge. The authors adopted a miniature artificial language learning set-up in which participants learned a language consisting of three words, which could be used to refer to two items. Crucially, two of the words were longer (seven versus three characters) than a third word. The two longer words could only be used to refer to one of the two objects. The third word, an abbreviation of the longer forms, could be used to refer to either of the two objects (it was ambiguous).\par
Operationalizing production cost as a function of word length, they hypothesized that when given a choice between alternatives, subjects should prefer to use the shorter, less costly term to refer to a more frequently occurring object. Importantly, however, this choice should only appear when there were both speaker- and listener-pressures present. Results indicated that speakers did tend to use the less costly form for the more frequent item when both pressures were present, but, importantly, not in conditions where only one or neither of the pressures were present.\par

While Kanwal et al. (2017) represents an important experimental step in confirming Zipfian notions about lexicon-level efficiency, the authors highlight an important and subtle dimension to these results. Zipf’s LOA is a lexicon-level theory. That is, it attempts to explain why particular word forms are mapped to particular meanings. However, their experimental results are, in theory, consistent with a pragmatic language-use account as well. That is, the underlying lexicon learned by subjects may have remained unchanged during the experiment -- a finding which, on its face, is at odds with the claim that this experiment demonstrates lexicon-level change. Rather, subjects may have *used* the lexicon pragmatically, preferring to map less costly forms to more frequent objects, while leaving the underlying lexicon unchanged. The authors’ highlight that because they only recorded participants’ actual language production they cannot distinguish between the lexicon-change account and pragmatic, language-use account.\par

The current project addresses the question of whether patterns of production behavior observed in Kanwal et al. (2017) can be explained by a pragmatic, language-use  account. To do so, we conduct a Bayesian data analysis, modeling subjects as rational pragmatic agents using the Rational Speech-Act framework (RSA) (Frank & Goodman, 2017; Goodman & Frank, 2016). To preview our results we find good fit from model posterior-predictive values to Kanwal et al. (2017) participant data  ($r^2=0.972$), over and above what would would be expected from a baseline model ($r^2=0.466$). We believe this provides compelling evidence in support of the alternative hypothesis outlined by Kanwal et al. (2017) -- that subjects may have been behaving pragmatically, rather than demonstrating real change at the level of language-structure. \par
The paper proceeds as follows. We describe the experimental set-up of Kanwal et al. (2017), their hypotheses and the alternative interpretation of their results. We then introduce our general approach to the Bayesian data analysis as well as the Rational Speech-act framework we will use to model subject’s pragmatic language use. We evaluate our results summarising the findings and highlighting dimensions of their analysis our current implementation does not yet capture. Following this work we return the question posed by Kanwal et al. (2017) of how pragmatic language use might bootstrap language change and offer an argument for how this type of process could in theory unfold.\par

>>>>

Zipf (1935) presented a view of human behavior that focused on *effort minimization* as a guiding principle. Relying on empirical evidence from language corpora, he argued that the unique dynamics of speaker- and listener-effort minimization give rise to the distributional forms found  in natural language. Among these, Zipf examined the relation between a word's frequency in a corpus and a variety of properties including its rank-frequency, its denotation size, and its length. In terms of this last property, what he called the "Law of Abbreviation," (LoA) Zipf claimed that the more frequent words are, the shorter they tend to be.\par

Zipf argued that this relationship between frequency and length was an emergent property of competing speaker-listener pressures what he called “Speaker’s economy” and “Auditor’s economy”. Subsequent work has argued that the LoA and related properties can theoretically be explained by a process of random typing (Ferrer-i-Cancho & Mascoso del Pardo, 2012). Kanwal et al. (2017) highlight that Zipf’s original work, as well as more recent studies analysing large-scale, observational corpora, “do not explicitly target the hypothesized role of communicative pressures…” In other words, it is difficult to make causal claims about the role that competing speaker-listener pressures play from observational data analysis.

To gain traction on the causal role of speaker-listener pressures, Kanwal et al. (2017) implemented an artificial language learning paradigm to isolate the pressures directly. In line with Zipf's original analysis, they hypothesized that only when *both* speaker- and listener- pressures are present in a communication setting would the LoA emerge. The authors adopted a miniature artificial language learning paradigm in which participants learned a language and completed a reference game task. The experiment included three conditions which directly manipulated speaker- and listener- pressures. They found that the language produced in the conditions were largely consistent with a Zipfian account -- only in the condition in which both speaker- and listener-pressures were present did production data follow LoA predictions.\par

While Kanwal et al. (2017) results are consistent with a Zipfian account of language change they highlight that their results are also consistent with a pragmatic, language-use account. That is, the results may reflect pragmatic use of an \textit{unchanging lexicon}. This observation is consistent with recent work which has begun to probe the question of how pragmatic processes, occuring in-the-moment of communication may give rise to long term effects in language -- perhaps such as those described by Zipf. Goodman & Frank (2016) highlight that understanding in an individual communication event can alter learning processes and lead to systematic changes in word meanings (Smith et al., 2013). Further, words can be pragmatically extended or narrowed in their denotations. The effect of such pragmatic use over time may give rise to longer-time lexical-level change.

Kanwal et al. (2017) give us an opportunity to explore the question of how long-term lexical change might be related to in-the-moment pragmatics. As pointed out by the authors, the experimental results are, in theory, consistent with both a Zipfian language-change account as well as pragmatic language-use account. Because they only recorded participants’ actual language production behavior, however, they cannot distinguish between these accounts using normal data-analytic methods. To do so would require access to the underlying \textit{mental representation} of the lexicon used by participants. The Rational Speech Act framework (RSA) (Frank & Goodman, 2012; Goodman & Frank, 2016) provides a useful computational formalism to gain traction on the question of pragmatics posed by Kanwal et al. (2017). RSA can largely be understood as a formulation of Gricean pragmatics, casting natural language understanding and production as probabilistic inference in agents who reason about one-another. \par

The current project compares a rational, pragmatic account as instantiated by RSA and a Zipfian, language-change account in which probabilistic agents use a modified lexicon with some error-probability. We conduct a Bayesian data analysis, inferring subject-level parameters and comparing posterior-predictive values for each participant’s data. To preview our results we find good fit from the pragmatics account, both overall ($r^2=XXX$), and in the crucial speaker- and listener-effort condition ($r^2=XXXX$); above what we find from the “Zipfian, language change” model ($r^2=XXXXX$).

We interpret these results in two ways -- first we believe it supports the importance of pragmatics in the particular study design of Kanwal et al. (2017). More generally, however, and in line with the authors’ discussion, we believe these results engage deeper questions on the relationship between pragmatics and language-change, in which pragmatica language use can be seen as an important first step in bootstrapping later change.\par

The paper proceeds as follows. We describe the experimental set-up of Kanwal et al. (2017), their hypotheses and the alternative interpretation of their results. We then introduce our general approach to the Bayesian data analysis as well as the computational instantiations of the two theories with discussion of the Rational Speech-act framework. We evaluate our results summarising the findings and highlighting dimensions of their analysis our current implementation does not yet capture. Following this work we return the question posed by Kanwal et al. (2017) of how pragmatic language use might bootstrap language change and offer an argument for how this type of process could in theory unfold.\par
